### 2020-06-04 report/sections/server/node/runtime.tex
Report: progress on chapter 4 runtime

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend {
	type: 'Send' = 'Send';
	abstract performSend(
		next: EfsmTransitionHandler,
		sendMessage: (role: Roles.Peers, label: string, payload: any[])
			=> void,
	): void;
};

@@

@@

\begin{itemize}
\item implementation details delegated to the actual implementation states
\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
\item visualise the interaction between runtime and states as a sequence diagram for message passing
\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
\end{itemize}

@@

@@

	performSend(
		next: EfsmTransitionHandler,
		sendMessage: (role: Roles.Peers, label: string, payload: any[])
			=> void
	) {
		const [label, payload, successor] = this.handler;

		sendMessage(Roles.Peers.Client, label, payload);
		return next(successor);
	}
}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodesend}
\end{figure}

\subparagraph{Sending through WebSockets}
Message Types are defined as interfaces, which
are represented by objects. By convention,
we serialise messages into \textit{JavaScript Object Notation}
(or JSON) \cite{json} using the built-in \texttt{JSON.stringify()}
method.

\begin{lstlisting}[language=javascript]
sendMessage(role: Roles.Peers, label: string, payload: any[]) {
	this.roleToSocket[role].send(JSON.stringify({
		label, payload
	});
}
\end{lstlisting}

@@

\subsection{Receiving Messages}
\label{subsection:noderuntimereceive}
\begin{itemize}
\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
\item motivate the edge case that message can arrive `out of protocol order' 
\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
\end{itemize}

@@

\subsection{Handling Termination}
WebSocket connections should be closed when the session terminates,
and both the browser endpoint and the server endpoint are capable
of closing connection. As we generate code for both, we define
a convention that the browser endpoint will initiate the WebSocket
close event.
As a result, we do nothing for this state at \cref{lst:noderuntme}.

>>>>

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend {
	type: 'Send' = 'Send';
	abstract performSend(
		next: EfsmTransitionHandler,
		send: (role: Roles.Peers, label: string, payload: any[]) 
			=> void,
	): void;
};

@@

\begin{remark}
We need to be careful when passing \textit{instance} methods
as \textit{function} arguments 
-- namely, the semantics of \lstonelinejs{this} is different.
In short, we have to \textit{explicitly}
\texttt{bind()} the \texttt{Session} object
to the instance methods that we pass as function arguments:

\begin{lstlisting}[language=javascript,tabsize=2]
// Inside Session class constructor...
this.next = this.next.bind(this);
this.send = this.send.bind(this);
this.register = this.register.bind(this);
\end{lstlisting}

Not doing so will result in \lstonelinejs{this} taking
a different value 
(either the global object or \lstonelinejs{undefined}).
\end{remark}

@@

@@

This is rather straightforward: 
we show the generated code in \cref{lst:nodesend}.

@@

	performSend(
		next: EfsmTransitionHandler,
		send: (role: Roles.Peers, label: string, payload: any[]) 
			=> void
	) {
		const [label, payload, successor] = this.handler;
		send(Roles.Peers.Client, label, payload); (*@\label{line:nodesendrole}@*)
		return next(successor);
	}
}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodesend}
\end{figure}

We get the label, payload and successor implementation
directly from the handler implemented by the developer, 
accurately typed by how we define the handler API in EFSM.ts.
The developer does not need to specify which role
to send to: this is a sensible design choice, as we know this
from the Scribble protocol, so we do not need the developer 
to specify separately.
As a result, we generate the code to send the message
to the correct role (\cref{line:nodesendrole}).
We use the \texttt{send()} method passed down
by the runtime to commit our communication action:
the runtime will handle how to serialise the message and perform
the send. We guarantee that \texttt{send()} is called
\textbf{exactly once} by construction, thus channel linearity
is never violated.
Finally, we use the parameterised EFSM transition handler
to notify the rutime which specific state to transition to.

\subparagraph{Sending through WebSockets}
We define message structures as interfaces, which
are represented by objects. 
By convention in \fancyname{SessionTS},
we serialise messages into \textit{JavaScript Object Notation}
(or JSON) \cite{json} using the built-in \texttt{JSON.stringify()}
method.

\begin{lstlisting}[language=javascript]
send(role: Roles.Peers, label: string, payload: any[]) {
	this.roleToSocket[role].send(JSON.stringify({
		label, payload
	});
}
\end{lstlisting}

@@

\subsection{Receiving Messages}
\label{subsection:noderuntimereceive}

We need to update the message event listener on the WebSocket
to use the developer's handler -- 
specific to the \textit{current state} -- 
to process the message.
Our approach is to keep the WebSocket message event listener
untouched, but define it in a way that allows \textit{dynamic} behaviour.
We walk through the concept implemented in (\cref{lst:noderuntimewsmsg}):

\begin{enumerate}
\item 
\texttt{Session} keeps track of the \textit{current}
message receive handler (\cref{line:noderuntimehandler}).
The \texttt{?} syntax denotes it is an \textit{optional}
type: not every state is a receive state, so there does not \textit{have} to
be an active message handler.

\item
The receive handler does \textit{not} need a specialised type
(\cref{line:msghandler}). The receive handler is defined in
the \texttt{Implementation} class of the concrete receive state,
so it will deserialise the message to the correct form.

\item
The \texttt{register()} method (\cref{line:register})
is passed to the \texttt{Implementation} class of the concrete
receive state, which will construct the message handler
around the developer's handler implementation and register it
with the runtime.

\item
When a message is received from the channel,
we dynamically process it with 
the current registered handler (\cref{line:callhandler}).
By MPST-safety, the incoming message must correspond
to the structure expected by the current receive state,
so correctness is preserved.
We encapsulate this dynamic behaviour in an instance method
and bind it as an event listener (\cref{line:wsbind})
for the WebSocket connection
of each non-server endpoint.
\end{enumerate}

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
type MessageHandler = (message: any) => void; (*@\label{line:msghandler}@*)

class Session {
	private handler?: MessageHandler; (*@\label{line:noderuntimehandler}@*)

	constructor(...) {
		...
		Object.values(this.roleToSocket).
			.forEach(ws => ws.onmessage = this.receive.bind(this)); (*@\label{line:wsbind}@*)
		this.handler = undefined;
	}

	register(handler: MessageHandler) { this.handler = handler; } (*@\label{line:register}@*)

	receive({ data }: WebSocketMessage) {
		const handler = this.handler;
		this.handler = undefined;
		handler?.(data); (*@\label{line:callhandler}@*)
	}
}
\end{lstlisting}
\captionof{lstlisting}{Approach to Dynamic WebSocket Message Event Listener}
\label{lst:noderuntimewsmsg}
\end{figure}

We also show the generated code for the \texttt{Implementation}
class of the receive state in \cref{lst:nodereceive}
-- this should appear consistent
with the explanation above.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
export class S51 extends IReceive {

	constructor(private handler: Handler.S51) { super(); }

	prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void
	) {
		const messageHandler = (message: any) => {
			const decoded = JSON.parse(message) as Message.S51;
			switch (decoded.label) {
				case Labels.S51.ADD: {
					const successor = 
						this.handler[decoded.label](...decoded.payload);
					return next(successor);
				}
				case Labels.S51.QUIT: {
					// Same as above...		
				}
			}            
		}
		register(Roles.Peers.Client, messageHandler);
	}
}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodereceive}
\end{figure}

Ideally, a more succinct (and direct) representation would be
\begin{lstlisting}[language=javascript]
const successor = this.handler[decoded.label](...decoded.payload);
return next(successor);
\end{lstlisting}

But this expresses a type dependency between \texttt{label}
and \texttt{payload}, which as discussed, cannot be implemented.
But message structures \textit{precisely} define a discriminated
union, so we handle this with a switch statement,
at the cost of having the same code in each case -- but
TypeScript does infer different types in each case statement,
so the duplicated code does serve a functional purpose.

Returning to \cref{lst:noderuntimewsmsg},
note that the type of the \texttt{handler} property is \textit{optional}
-- this hints at a consistency problem: 
\textit{how do we know that \texttt{this.handler} is set when
a message is received?}

\begin{itemize}
\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
\item motivate the edge case that message can arrive `out of protocol order' 
\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
\end{itemize}

@@

\subsection{Handling Termination}
WebSocket connections should be closed when the session terminates,
and both the browser endpoint and the server endpoint are capable
of closing connection. As we generate code for both, we define
a convention that the browser endpoint will initiate the WebSocket
close event.
As a result, we do nothing for this state at \cref{lst:noderuntime}.
